---
title: "The 'freezing hypothesis' revisited"

subtitle: "Session 06"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

navigation: horizontal 

output:

  beamer_presentation:

    #includes:
    #  in_header: header_pagenrs.tex

    latex_engine: xelatex
    
    theme: "Pittsburgh"
      
    #fonttheme: "serif"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: TRUE
    
    toc: TRUE
    
    number_sections: TRUE

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files

# Loading images' paths
figure1_chiaramonte17 <- "images/figure1_chiaramonte17.png"
figure4_chiaramonte17 <- "images/figure4_chiaramonte17.png"
figure5_4_bertoa21 <- "images/figure5_4_bertoa21.png"
```


# Overview of the last session

## Overview of the last session

Take-home points from Session 5:

  1. **Spatial models of voting and party competition** are useful for understanding party systems dynamics. They conceptualize elections as markets in which **voters and parties have preferences over policy dimensions that reflect lines of conflict**
  
  2. The main dimension of competition is the **left-right dimension**, that represents preferences for distributional policies deeply rooted in the **class cleavage**
  
  3. Since this dimension cannot account for major changes in Western European party systems, many scholars agree that a new dimension representing **a new cultural cleavage is re-structuring political competition**
  
  4. Hence, **political competition in Europe is better understood within a two dimensional political space**
  
# Assesing party-system change

## Assesing party-system change

- We have discussed a lot of ways of looking at change and continuity already

- **What is your impression so far?**

- Well, summarising all these aspects is difficult indeed

- That's why some scholars have proposed different measures that synthesize the different features that define party system stability

- In this session, we will review three major contributions:

  1. The contribution of **Mair** (RIP) and Bartolini [@Bartolini1990; @Mair1993]
  
  2. The contribution of Chiaramonte and Emanuele [@Chiaramonte2017; @Emanuele2019; @Emanuele2018]
   
  3. The contribution of Bértoa and Weber [@CasalBertoa2019]

# The freezing hypothesis revisited

## The freezing hypothesis revisited

- Some context on the freezing hypothesis in the 1990s:

  1. The work of Lipset and Rokkan [-@Lipset1967a] on the formation of party systems and the **cleavage structure in Europe** remained **largely uncontested**
  
  2. However, there was a **huge criticism towards the freezing hypothesis** since the 1970s
  
  3. As we advanced in *session 3*, the **rise of new parties** plus patterns of **weakening party ties** and **higher electoral volatility** led many scholars to claim that party systems had indeed transformed

## The freezing hypothesis revisited

- The general argument was:

"It is not so much that Rokkan was wrong; rather, what has happened is simply that his analysis has been overtaken by events" [@Mair1993, p. 122]

- It is in this context that Peter Mair and Stefano Bartolini carry out a systematic comparative analysis and set the basis for analyzing party system change

## The freezing hypothesis revisited

- They propose to assess three claims:

  1. Whether there is a pattern of **de-alignment** (the decline of party)
  
  2. Whether there is a pattern of **re-alignment** (the emergence of new parties)
  
  3. Whether there is general **electoral instability** (both)
  
- Using two empirical indicators:

  1. Electoral volatility
  
  2. Traditional party support

- Is party system change a "myth"? [@Mair1993, p. 123]

## The freezing hypothesis revisited

1. Electoral volatility

    - It can range from 0 (no change) to 100 (total change) [@Pedersen1979]
  
    - Bartolini and Mair [-@Bartolini1990] analyse 303 elections in the period 1885-1985 [-@Bartolini1990]

    - **Average level between 1945 and 1989**: 8.7 (meaning 91.3 in stability)
  
    - Some exceptions: Germany in 1920 (32.1) OR France in the 1950s (22.3 in average)
  
    - Importantly **higher in the inter-war period** (in which the frozen hypothesis is based on): 9.9

  
## The freezing hypothesis revisited

1. Electoral volatility

    - "In fact, if one looks at the three broad electoral epochs which have occurred since universal suffrage and the age of mass politics - the inter-war period, the first post-war decades, and then the most recent post-war decades, **aggregate electoral volatility, aggregate electoral instability, has shown a consistent decline**" [@Mair1993, p. 124] 

    - In addition, **most volatility happens within blocks**
    
    - Average **cross-cleavage volatility** in the post-war and the interwar periods: 2.9 and 3.2 respectively
    
    - Still, some differences across European countries, **although minor**
    
## The freezing hypothesis revisited

2. Survival of traditional parties

    - To assess de-alignment rather than general instability

    - Comparing electoral support for traditional parties from the early 60s ('frozen period') to the late 80s and early 90s ('changing period')
    
    - **95% of the vote in average in the 'frozen period'** compared to **84% of the vote in the 'changing period'**
    
    - Also in the upper and lower bounds, **10% decrease**
    
    - **Is this a lot or not?**
    
    - According to Mair, **it is not**
    
    - Particularly relevant given the generational replacement and the enfranchisement of new voters

## The freezing hypothesis revisited

- "(...) **is the cleavage system or the party system frozen?** If we say that the cleavage system is frozen, we are saying that the social groups bounded by a cleavage will support the same party generation after generation." (Andrea's response paper)

- "A frozen party system is a structurally consolidated or institutionalised system, but **if society changes**, it leads me to ask **how do parties manage to maintain or even strengthen their links with social groups?**" (Andrea's response paper)

## The freezing hypothesis revisited

- In sum, no obvious change

- But why party systems are bias towards stability while political analyses are bias towards instability?

  1. The appeal of newness
  
  2. **Misconception of the social structure-political structure interaction** [e.g., @Franklin1992]
  
      - "parties as independent actors capable of shaping the environments in which they compete to survive" (Andrea's response paper)
  
  3. Underestimation of the capacity of parties to adapt

- **But is the evidence summarised by Mair enough?**

# Twenty years later: new evidence from a longer perspective

## Twenty years later: new evidence from a longer perspective

- In the past few years, the debate on party system change has revitalized due to

  1. dramatic volatility in SE
  
  2. the rise of PRR parties in former right-wing authocracies (e.g., Germany)

- Chiaramonte and Emanuele provide a series of studies updating the analyses of Mair. **Two approaches**:

  1. **Party system innovation** [@Emanuele2018; @Emanuele2019]

  2. **Volatility and its components** (**Type A** [from old to new parties] vs. **Type B** volatility [among old parties]) plus **sequential volatility** to  assess **de-institutionalization** [@Chiaramonte2017] \footnotemark
  
    \footnotemark[1]Recall that we defined **party system institutionalization** as the degree in which parties act in a predictable and patterned way [@CasalBertoa2019b, p. 569]
    
## Twenty years later: new evidence from a longer perspective

```{r, fig.show="hold", out.height = "60%", out.width="100%", fig.cap = "Figure 1 from Chiaramonte and Emanuele (2017)"}
include_graphics(figure1_chiaramonte17)
```  

## Twenty years later: new evidence from a longer perspective

- The results:

  1. **Party system innovation** has grown, specially since 2010

  2. **Electoral volatility and is components**: 
  
      + "the overall picture of the levels of party system volatility and regeneration across time and space shows that the **‘fundamental bias towards stability’** emphasized by Bartolini and Mair (1990: 68) **appears confirmed**"
      
      + **however**, "there is a clear increase of both variables during the last electoral period (1992–2015), with further acceleration since 2010 that reveals **a general shift towards instability in the last years**"

## Twenty years later: new evidence from a longer perspective
      
```{r, fig.show="hold", out.height = "60%", out.width="80%", fig.cap = "Figure 4 from Chiaramonte and Emanuele (2017)"}
include_graphics(figure4_chiaramonte17)
```  

## Twenty years later: new evidence from a longer perspective

- Another long-term approach is provided by Bertoa [-@CasalBertoa2019]
  
- The theory of **restrained change**:

  1. **Economic crises impact party systems asymetrically**
  
  2. They tend to **disturb party systems that are generally restrained** 
  
  3. In contrast, they **consolidate party systems that are already complex**
  
  4. **That's why party systems rarely fall apart of ultimately de-institutionalize**
  
- Mechanism: **risk aversion** in both parties and voters

- Empirical patterns confirmed in the **1929, 1973 and 2008 crises**

- Exceptions: **Italy in 1994?**

## Twenty years later: new evidence from a longer perspective

- How to measure **party system restraint**:

  1. Electoral volatility
  
  2. **Party system fragmentation** (i.e., effective number of parties: weighted number of parties either with parliamentary presence or passing an arbitrary electoral threshold)
  
  3. **Polarization** (i.e., weighted ideological distance between the effective parties)
  
  4. **Party system closure** (i.e., the degree to which govt. formation allows innovation: [1] alternation in govt., [2] government formulas stability [3] openess to new parties for govt. formation) [@Bertoa2021]

## Twenty years later: new evidence from a longer perspective

```{r, fig.show="hold", out.height = "60%", out.width="80%", fig.cap = "Figure 5.4 from Bertoa and Zsolt (2021)"}
include_graphics(figure5_4_bertoa21)
```  

# Measuring party-system change

## Measuring party-system change

- A **summary of measures** to assess party system change:

  1. Electoral volatility (cross-cleavage vs. between-cleavage/type A vs type B/electoral sequences)
  
  2. Traditional parties strength
  
  3. Party identification/membership [@Mair2013]
  
  4. Party system innovation
  
  5. Party system fragmentation
  
  6. Polarization
  
  7. Party system closure
  
  8. **Conflict dimensionality**? [based on @Hutter2019]

## Measuring party-system change

- One controversial arguments in Mair [-@Mair1993] is that **even if the social alignment of parties have changed, this doesn't mean that cleavage politics is dead**. This argument doesn't consider whether ***cleavage politics has changed***

- ***Final activity!*** Let's divide the class in two groups. You have 5-10 minutes to think of new indicator to assess party system change. Then, you will have to explain why this indicator and defend it while trying to (constructively) criticise the indicator proposed by the other group.

## References {.allowframebreaks}

<div id="refs"></div>
