---
title: "New far right parties and the (silent) counter-revolution"

subtitle: "Session 03"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

navigation: horizontal 

header-includes:
- \usepackage{booktabs}

output:

  beamer_presentation:

    #includes:
    #  in_header: header_pagenrs.tex

    latex_engine: xelatex
    
    theme: "Pittsburgh"
      
    #fonttheme: "serif"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: TRUE
    
    toc: TRUE
    
    number_sections: TRUE

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files
library(dagitty) # for DAGs

# Loading images' paths
ignazi <- "images/table1_ignazi.png"
norris <- "images/norris_inglehart.jpg"
mudde <- "images/table1_mudde.png"
```


# Overview of the last session

## Overview of the last session

Take-home points from Session 2:

  1. **Party-systems** are strongly **embedded** within the institutions and the socio-economic structure of countries. They are **path dependent and difficult to change**
  
  2. However, **changes in the socio-economic structure** accompanied by **mobilization trough extra institutional channels** can generate an **opportunity window** to newcomers
  
  3. These **changes**, however, **are unlikely to be large**, because **the conflict structure tends to absorb new issues**

## Overview of the last session

  4. This has been the case of the **green parties** so far. They were born as a response to societal changes (**the silent revolution**) and the mobilization of new issues (**new social movements**) and **became institutionalized in many cases but with limited success**
  
  5. Their success depended on (1) the degree of social change, (2) the degree of institutional constraint, (3) the openness of the party system and (4) and the structure of political competition
  
  - In this session we will discuss how **change can trigger change**; and that **political dynamics are not often linear** 
  
# The weakening of party ties

## The weakening of party ties

- In the 1980s, there was a growing concern that **party ties** (i.e., the link between society and the political parties) were **fading away**

- The **number of party members and party identifiers** was at a low peak 

- The silent revolution and **the growing important of green and new left parties** made many scholars think that rooted political identities were weakening

- Instead, people seemed to vote based on their individual preferences on temporarily salient issues rather than as an expression of group identity (i.e., **the 'issue voter'** thesis)

## The weakening of party ties

- It also gained importance the thesis that established parties, because of the weakening of party ties, were increasingly shifting towards a **'catch-all' model** [@Kirchheimer1966], i.e. party competition as policy markets

- Instead of defending the interest of a specific part of the electorate, they would **moderate or blur their positions on controversial issues to attract a wider audience**

- Thus, they would devote **less effort to develop programmatic linkages and a a strong organization**

- Instead, their mobilization strategies would increasingly focus on altering the **saliency** of **valence issues** in which to prove their credibility as capable managers of the *res publica*

## The weakening of party ties

```{r echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Dynamics of social and electoral change"}

dag <- dagitty("dag {
  PARTY_MEMBERSHIP_DECLINE -> WEAKENING_PARTY_TIES
  PARTY_IDENTIFICATION_DECLINE -> WEAKENING_PARTY_TIES
  WEAKENING_PARTY_TIES -> ISSUE_VOTING
  WEAKENING_PARTY_TIES -> CATCH_ALL_STRATEGIES
  CATCH_ALL_STRATEGIES -> ISSUE_VOTING
               }")

coordinates(dag) <-
    list(x = c(PARTY_MEMBERSHIP_DECLINE=5,
              PARTY_IDENTIFICATION_DECLINE=5,
              WEAKENING_PARTY_TIES=4,
              CATCH_ALL_STRATEGIES=2.5,
              ISSUE_VOTING=3.25),
        y = c(PARTY_MEMBERSHIP_DECLINE=3.25,
              PARTY_IDENTIFICATION_DECLINE=2.75,
              WEAKENING_PARTY_TIES=3,
              CATCH_ALL_STRATEGIES=3,
              ISSUE_VOTING=2.625))
          

plot(dag)

```

## The weakening of party ties

- This would lead to three main outcomes:

  1. A higher **electoral volatility** (i.e., net vote changes between one election and the following)
  
  2. The rise of **new parties**
  
  3. The **decline of parties as such** (i.e., socially rooted organizations)

- But **were these outcomes true?**

- Let's focus on focus on the first two!

## The weakening of party ties

- Very **few parties had succeeded** in passing the threshold of representation

- Among the few that had done it, **many disappeared** or fell back into a marginal role

- Changes at the societal and partisan level had **not undermined** the **cleavage structure**, i.e. votes remained overwhelmingly within each political family

- "As convincingly argued by Bartolini and Mair [-@Bartolini1990], the long-awaited ‘un-freezing’ of partisan alternatives has yet to come." [@Ignazi1992, p. 5]

- In other words, the undergoing process seemed to one of ***de-alignment*** **rather than** ***re-alignment***

# The (silent) counter revolution

## The (silent) counter revolution

- However, some electoral changes beyond the timid advances of the green parties and teh new left were noticed in the end of the 80s and the beginning of the 90s

- N**ew extreme right parties**, different from conservative and liberal parties had gained momentum in some national elections (remarkably, the National Front in France)

- Ignazi [-@Ignazi1992] was the first to call the scholarly attention to this phenomenon

- Most importantly, these parties **resembled those extreme right parties of the inter-war period because they were able to attract highly diversified voters**, cross-cutting existing cleavages

- On the other hand, they were different from the inter-war extreme right parties and their **successors** because **they didn't reject the democratic principles of popular sovereignty**

## The (silent) counter revolution


```{r, fig.show="hold", out.height = "85%", out.width="85%", fig.cap = "Table 1 from Ignazi (1992)"}
include_graphics(ignazi)
```


## The (silent) counter revolution

- A new question emerges then: **"Why, in an era of mounting postmaterialism and economic growth, do we find an increasing number of rightwing voters?"** [@Ignazi1992, p. 5]

- It is important to notice the formulation of this question thirty years ago because hundreds of papers are still published every year trying to answer it

- Let's reflect on the answer given by Ignazi: that as the same way that a silent revolution had occurred, a (also) **silent counter-revolution** was struggling to be born

## The (silent) counter revolution

- Now let's take a look at the cover of a political science best-seller book published in 2019 with more than two thousand citations in  February 2022 

```{r, fig.show="hold", out.height = "60%", out.width="60%", fig.cap = "Cover of Cultural Backlash (2019)"}
include_graphics(norris)
```

## The (silent) counter revolution

- The silent counter-revolution thesis:

  1. A renewal importance of **traditional values** as a response to the threat of post-materialistic values
  
  2. Increasing ideological **polarization**
  
  3. A **decline of system legitimacy**

  4. "Immigration, law and order." The rise of a **new value dimension**
  
- **We will talk extensively about the last point tomorrow!**

# A new party family? The populist radical right

## A new party family? The populist radical right

- Another important take on the new extreme parties come from Beck [@Betz1993]

- He made two important contributions:

  1. Labelling this growing party family as the **populist radical right** (PRR)
  
  2. Coined the term of **political resentment** to explain its success

## A new party family? The populist radical right

- Some scholars pointed to **the post-industrial transition** as one of the main factors explaining support for PRR parties, as unskilled and semi-skilled workers had to compete with the incoming immigrant population for increasingly underpaid jobs (*economic explanation*)

- **The politics of resentment** introduce an alternative and complementary explanation. These people would be also disappointed by the promises of growth made during their youth, and the aid to foreign-born population and the new middle classes provided by the public service (*cultural explanation* and counterpart to the silent revolution thesis)

## A new party family? The populist radical right
  
- On the **PRR** definition:

  1. They are "are **radical** in their rejection of the established sociocultural and sociopolitical system and their advocacy of individual achievement, a free marketplace, and a drastic reduction of the role of th"
  
  2. "They are **right-wing** in their rejection of individual and social equality, in their opposition to the social integration of marginalized groups, and in their appeal to xenophobia, if not over"

  3. "They are **populist** in their instrumentalization of sentiments of anxiety and disenchantment and their appeal to the common man and his allegedly superior common"

## A new party family? The populist radical right

"In short, they tend to combine a **classic liberal position on the individual and the economy** with the **sociopolitical agenda of the extreme** and intellectual **new right**, and they deliver this amalgam **to those disenchanted with their individual life chances and the political system.**" [@Betz1993, pp. 413-4]

## A new party family? The populist radical right

- Today, Mudde [-@Mudde2014] talk of three defining features:

  1. **Nativisim**: "a xenophobic form of nationalism in which a mono-cultural nation-state is the ideal and all non-natives (i.e. aliens) are perceived as a threat to the nation"
  
  2. **Authoritarianism**: "a strict belief in order and its stringent enforcement within society through discipline, law and order-based policies"
  
  3. **Populism**: "a thin ideology that considers society to be essentially divided between two antagonistic and homogeneous groups, the pure people and the corrupt elite, and wants pol- itics to reflect the general will ofthe people"

- **A new party family?**

# The impact of PRR across Europe

## Comparing first movers: Austria & France

- We can distinguish two types of PRR parties, according to their origins: 

  1. **Converted old parties**, like the FPÖ in Austria or the SVP in Switzerland

  2. **Truly new parties**, like the FrP in Denmark and the FN in France

## Comparing first movers: Austria & France

- The **Austrian FPÖ** was a liberal conservative party that had lost support in the mid 1980s but recovered force after the young charismatic and populist Jörg Haider was elected to the chair of the party in 1986. In 1991 became the second largest party at the national level. In 1999 became the most voted party in a general election and formed a coalition government with the Christian-democrat ÖVP

- The **French Front National** was founded in 1972 by right-wing radicals and gained from virtually zero in the 1981 elections to 9.6% of the vote in 1988. However, it wasn't until a moderation shift by Marine Le Pen in 2011 that the party became the third largest party in a presidential election, and then second the largest party in 2017

## The impact of PRR across Europe


```{r, fig.show="hold", out.height = "75%", out.width="75%", fig.cap = "Table 1 from Mudde (2014)"}
include_graphics(mudde)
```

## Current trends on the PRR

- At least three new big events has changed the picture displayed in the previous slide:

  1. **The Great Recession** (i.e., the 2008 financial crisis and its consequences)
  
  2. **The 2015 refugee crisis**
  
  3. **Brexit**
  
- This have lead to rapid changes:

  1. The emergence and success of **AfD in Germany**
  
  2. The emergence and success (but also fast decadence and ostracism) of **UKIP/Brexit Party in UK**

  3. The end of the 'Iberian exceptionalism' (**Vox and Chega in Spain and Portugal**, respectively)
  
- Here, there may play a factor that we haven't discussed until now: ***diffusion and international contagion***

## The impact of PRR across Europe

- What does this picture say about what we have discussed until now?

- **Let's debate!** This time, I want you to think alone about the factors that affect party system stability (change) and that can be summarised, as we saw in Session 2, in the points below. You have 5 minutes to think about how each of them may has created barriers/opportunities to the success of RPP parties.Then, we will go one by one to debate your arguments together

  1. **Societal factors** (stability of conflicts/cleavages and the electorate)
  2. **Institutional factors** (crucially electoral rules)
  3. **Party-system factors** (existent alliances, within-block fragmentation, etc.)
  4. **Party-level factors** (party linkages, strategies and organization)
  5. **Individual-level factors** (habits, attitudes and group identities)


# References

## References {.allowframebreaks}

<div id="refs"></div>
