---
title: "University of Lucerne<p> </p>**Comparative Western European Party Systems: Continuity and Change**"

date: | 
 Spring Term 2021-2022^[This version of the syllabus: `r format(Sys.time(), '%d %b %Y')`]

author: 
- Álvaro Canalejo-Molero^[PhD Candidate, Dept. of Political and Social Sciences, European University Institute, [Alvaro.Canalejo@EUI.eu](mailto:alvaro.canalejo@eui.eu)]


bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue

header-includes:
  
- \usepackage{setspace}\onehalfspacing
- \usepackage[utf8]{inputenc}
- \usepackage{dcolumn}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}

fontsize: 12pt

output: 
  
  pdf_document:
    
     latex_engine: xelatex
     
     extra_dependencies: ["flafter"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Installing packages
#install.packages("stevetemplates")
#install.packages("bibtex")
#install.packages("RefManageR")

# Setting bibliography settings
library(RefManageR)
bib <- ReadBib("../00_bibliography/cweps_syllabus.bib") # assign .bib file
myopts <- BibOptions(bib.style = "authoryear", style="latex", first.inits=FALSE, max.names = 20)
```


# Course description

Elections in Europe has become increasingly unpredictable in the last two decades. The number of parliament parties is growing, and coalitions are complicated to foresee. In Germany, for example, the duration of negotiations to form the government escalated from one month in 2013 to almost six months in 2017, after the entry of the radical right party AfD in parliament. In other European countries, such as Spain or the Netherlands, the number of parties in parliament has almost doubled in twenty years. As a result, anticipated elections and broken coalitions are now a permanent risk. 

How has this situation come about? These changes are not random but driven by societal transformations and the strategies of new political actors, like social movements and challenger parties. In addition, they have not affected all Europe equally but depend on contextual factors that vary across countries and over time.

This seminar will review the different 'transformation waves' that have shaped Western European party systems from the 1960s until now. We will then make sense of these changes using the most relevant (electoral) demand and supply factors discussed in the literature. The goal is to provide students with conceptual and empirical tools to analyse the evolution of Western European party systems from a historical and comparative perspective. By the end of the course, they will be able to identify patterns of continuity and change and critically assess future electoral scenarios. 

# Course organization

The course is organized as a four-day intensive seminar divided in two weekend blocks. The first block consists of a review of European party systems, focusing on their main changes from the 1960s until now. It is structured as follows. First, we will review the main arguments of the cleavage theory as originally proposed by Lipset and Rokkan (1967), making particular emphasis on the so-called 'freezing hypothesis' (i.e., European party systems in the 1960s largely resemble their original structure in the 1920s). Second, we will discuss the incorporation of two new party families into the political landscape in the following decades (Green and Radical Right Parties). We will assess the impact of these changes using the same indicators discussed by Mair (1993) and we will try to make sense of them by incorporating a cultural/value-based cleavage to the original cleavage theory. Finally, we will discuss the impact of the Great Recession on increasing electoral volatility and we will carry an empirical exercise with real-world data to inform a wrapping-up discussion. 

The second block discusses the most relevant demand and supply factors provided by the literature to explain party-system change. We will focus on changes in the socio-economic structure of modern European societies, such as the emergence of new educational divides, the formation of new social classes and the effect of globalization and immigration on altering the preferences of the electorate. On the supply side, we will briefly discuss the role of ideological convergence (among dominant parties) and issue entrepreneurship (by challenger parties) on carrying political change.

# Learning outcomes

By the end of this course, students will be able to:

- Identify the main theories of party system formation and change in Western Europe

- Apply these theories to analyse specific cases from a comparative and historical perspective

- Understand the concept of electoral volatility and interpret it using its most common empirical indicators

- Assess elections as markets in which demand and supply factors interact to produce electoral outcomes (i.e., seats and govt. coalitions)

- Read scientific articles and present critical written responses

- Work in collaboration with others

- Orally present their work to a broad audience

# Requirements and evaluation I (4 credits)

## Attendance and class participation

For obtaining the credits (pass), students are expected to attend at least 80% of the classes.

During class, we will carry interactive exercises, such as in-group debates or interpreting real-world data to derive hypotheses to the phenomena under study. Active engagement in those activities as well as interventions linking the content of the readings to them are expected.

## Readings and response papers

A set of readings for each session will be delivered through OLAT. Students are expected to complete all the mandatory readings (one per session), select one for each weekend (two in total) and write a short response paper (300-500 words) to them. The response papers should be submitted at least 5 days before the corresponding session to alvaro.canalejo@eui.eu.

## Class presentation

Students will be divided into pairs in the last part of the first block. They will be randomly assigned a Western European country (other than Switzerland) and will have to prepare a presentation (30-35 min.) in which they will describe the evolution of its party system until now. Students should make emphasis on periods of change (as opposed to continuity) and use relevant academic literature to justify their arguments. They are not expected to present simple historical reviews, but will be asked to incorporate some of the theories described in class to hypothesize the reasons why each party system has followed a specific pattern, and frame the case comparatively within the European context. Showing empirical evidence will not be mandatory but supporting their arguments with basic statistical analysis will be positively evaluated.

The presentations should be supported by visual material (power point slides or a similar format) and pose two or three questions at the end for class discussion.

# Requirements and evaluation II (4 credits for BA students / 6 credits for MA students)

## Seminar paper

Students choosing to write a seminar paper (around 15 pages for BA students and 25 pages for MA students) will investigate one of the course topic in greater depth. Please, if you are interested in doing a seminar paper for this class get in contact at alvaro.canalejo@eui.eu in advance to discuss the details.

# Course schedule

## Block I - Friday 04.03.2022 (09:00 – 17:00)


**Session 1 (09:00 – 10:30): Party-system formation in Europe and the 'freezing hypothesis'**

  + Required readings:

```{r, echo = FALSE, results="asis"} 
bib["Boix2007"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
#bib["Lipset1967a"]
bib["Lipset1967a", "Kalyvas2018"]
``` 

```{r, echo = FALSE, results="asis"} 
#bib["Lipset1967a"]
bib["Przeworski1986", "Merkl1969"]
``` 



**Session 2 (10:45 – 12:15): 'Green parties' and new social movements (70s-80s)**

  + Required readings:
    
```{r, echo = FALSE, results="asis"} 
bib["MullerRommel1998"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
bib["Inglehart1977", "Kriesi1989"]
``` 

```{r, echo = FALSE, results="asis"} 
bib["Kriesi1992", "Muller-Rommel2019"]
``` 



**Session 3 (13:45 – 15:15): The emergence of new 'radical right parties' (80s-90s)**

  + Required readings:
    
```{r, echo = FALSE, results="asis"} 
bib["Ignazi1992"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
bib["Mudde2007", "Mudde1999"]
``` 



**Session 4 (15:30 – 17:00): The 'freezing hypothesis' revisited: Mair & Bartolini in the 90s**

  + Required readings:
  
```{r, echo = FALSE, results="asis"} 
bib["Mair1993"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
bib["Bartolini1990"]
``` 

## Block I - Saturday 05.03.2022 (09:00 – 15:15)



**Session 5 (09:00 – 10:30): The cultural dimension: cleavage theory revisited**

  + Required readings:
    
```{r, echo = FALSE, results="asis"} 
bib["Bornschier2010"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
bib["Hooghe2018b", "Downs1957"]
``` 

```{r, echo = FALSE, results="asis"} 
bib["Kriesi1998", "Kriesi2006"]
``` 



**Session 6 (10:45 – 12:15): Economic voting and the Great Recession: North-Western and Southern European countries compared**

  + Required readings:
    
```{r, echo = FALSE, results="asis"} 
bib["Kriesi2019"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
bib["Hutter2018", "Hernandez2016"]
``` 



**Session 7 (13:45 – 15:15): Changing party-systems? Wrapping-up workshop on electoral volatility**

  + Required readings:
    
```{r, echo = FALSE, results="asis"} 
bib["Chiaramonte2017"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
bib["Emanuele2018","Emanuele2019"]
``` 

## Block II - Friday 08.04.2022 (09:00 – 17:00)



**Session 8 (09:00 – 10:30): Factors of change I: demand-side explanations**

  + Required readings:
    
```{r, echo = FALSE, results="asis"} 
bib["Oesch2008"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
bib["Cavaille2019"]
``` 

```{r, echo = FALSE, results="asis"} 
bib["Oesch2018","Rooduijn2018"]
``` 



**Session 9 (10:45 – 12:15): Factors of change II: supply-side explanations**

  + Required readings:
    
```{r, echo = FALSE, results="asis"} 
bib["Hobolt2012"]
``` 
      
  + Additional resources:

```{r, echo = FALSE, results="asis"} 
bib["Hobolt2015", "Hobolt2016"]
``` 

```{r, echo = FALSE, results="asis"} 
bib["Spoon2019"]
``` 



**Session 10 (13:45 – 15:15): Students presentations**



**Session 11 (15:30 – 17:00): Students presentations**


## Block II - Saturday 09.04.2022 (09:00 – 15:15)



**Session 12 (09:00 – 10:30): Students presentations**



**Session 13 (10:45 – 12:15): Students presentations**



**Session 14 (13:45 – 15:15): Wrapping-up discussion, advanced topics and feedback**

  + Recommended readings:

```{r, echo = FALSE, results="asis"} 
bib["Mair2013"]
``` 