---
title: "Factors of change II: supply-side explanations"

subtitle: "Session 12"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

navigation: horizontal 

header-includes:
- \usepackage{booktabs}

output:

  beamer_presentation:

    #includes:
    #  in_header: header_pagenrs.tex

    latex_engine: xelatex
    
    theme: "Pittsburgh"
      
    #fonttheme: "serif"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: TRUE
    
    toc: TRUE
    
    number_sections: TRUE


---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files

# Loading images' paths
fig1 <- "images/Hobolt_2012_fig1.png"
tbl1 <- "images/Hobolt_2012_tbl1.png"
tbl3 <- "images/Hobolt_2012_tbl3.png"
```

# Explaining party system change: demand vs. supply factors

## Explaining party system change: demand vs. supply factors

- From analyses at the **aggregate**-level (country, party-system or election level) to analyses at the **party and voter** (i.e., individual) levels

- We can distinguish between **supply** and **demand** side explanations
  
- The demand side (yesterday):

  + Level of analysis: individuals/voters
  
  + Question: who votes for new parties and why?
  
- The supply side (today):

  + Level of analysis: parties
  
  + Question: how do new parties get votes?

# Factors of change II: supply-side explanations

## Factors of change II: supply-side explanations

- Shifting from the analysis of countries, party-systems and elections to **parties** (not only: meso-level units of observations too [e.g., party families or types])

- **What supply-side factors may explain that some new/challenger parties are more successful than others?** (think on implications of  the theories we have reviewed!)

  + Charismatic leaders
  
  + Grassroot mobilization
  
  + ...

  + **Politization of new issues**

- ***When dimensions collide: The electoral success of issue entrepreneurs*** [@Hobolt2012]

# Issue politization: "The electoral success of issue entrepreneurs"

## The puzzle and research question

- Recent debates on the multi-dimensionality of political competition in the West (1 vs. 2 dimensions)

- **But how do dimensions change?**

- The theory of issue evolution in the US describes one strategy followed by **opposition parties** to become winners consisting of:

  1. **Polarizing** a previously non-salient issue
  
  2. Making this issue more **salience**

- But this theory assumes a two-party system, so ***Research Question***: **does this model** ***or*** **how does this model apply to multi-party systems like those in Europe?**

## The puzzle and research question

- This paper applies the theory of **issue evolution** to European multi-party system to argue that **new/challenger parties** are those driving dimensional change (*note the distinct emphasis here:* ***from voters to parties***)

- **QUESTION(s)**: what is the puzzle of this paper? do you remember the dimensionality debate?

## Hypotheses

- The issue evolution theory requires (assumptions!):

  + **Clarity**: voters must be aware of the different position of the parties on the new issue
  
  + **Affect**: voters must also care about this issue and the differences in party positions
  
- **QUESTION**: what parties are more likely to promote a new issue and why?

- "(...) losers on the dominant dimensions of competition have the most powerful incentive to promote a new issue" (similar to Riker's tehory of issue manipulation - similar to agenda setting) (p. 249)

## Hypotheses

```{r, fig.show="hold", out.height = "60%", out.width="80%", fig.cap = "Figure 1 in Hobolt and De Vries (2012)"}
include_graphics(fig1)
```

## Hypotheses

- The issue entrepreneurship strategy:

  1. Polarization 
  
  2. Centrality
  
- Both strategies together are aimed to increase the saliency of a new issue

- But what parties are more likely to carry out this strategy in a multi-party system? Three party categories:

  1. Mainstream government
  
  2. Mainstream opposition
  
  3. Challenger parties (**QUESTION**: What is a challenger party?)

- **QUESTION:** Why challenger parties are more likely to carry out strategies of issue entreprenurhsip than mainstream opposition parties in multi-party systems?

## Hypotheses

- The issue under study here: **European integration** (**QUESTION:** why?)

- "In previous work, we have shown that, when it comes to the issue of European integration, challenger parties are more likely issue entrepreneurs than mainstream parties" (p. 252)

- "Since European integration was conceived as a top-down project based on a broad elite consensus, the ‘mainstream’ position among European political parties has been broadly pro-integrationist" (p. 252)

## Hypotheses

- But is issue entrepeurhsip successful (i.e., does it alter the dimensions of competition)?

H1: Voters choosing to vote for a challenger party are more likely to base their vote choice on preferences related to the new issue dimension, all other things being equal.

H2: Parties engaging in an issue entrepreneurial strategy are more likely to increase their vote share, all other things being equal.

## Research design

- Study 1:

  - 21 West and East European countries in the year 2004
  
  - Individual-level survey data and expert-based data on parties
  
- Study 2: 

  - Time-series cross-sectional data of electoral results at the national level
  
  - With data expert-based data on party positions
  
  - Only for Western European countries
  
- **QUESTION:** what do you think of this research design?

## Data and models

- Study 1:

  - DATA: EES 2004 data on vote choice, evaluations of the EU (comparable surveys in member states) + CHES data for the parties' classification
  
  - MODEL: evaluations of the EU on vote choice + controlling for socio-economic characteristics, economic and government evaluations, left/right, post-communist legacy (multinomial logistic hierarchical regression model)
  
  - **QUESTION:** why these control variables?
  
  - "The socioeconomic controls include education, religiosity, and income. These latter variables are incorporated to control for dominant models explaining vote choice, such as economic and cleavage-based voting" (p. 255)
  
  - **QUESTION:** why the multilevel approach?
  
## Data and models

- Study 2:

  - DATA: electoral records from 1984 to 2006 + CHES data on issue saliency and position (in particular for position on European integration)
  
  - To capture ISSUE ENTREPRENEURSHIP: multiplying party salience on the issue by the party position on the issue minus the mean party position on the issue
  
  - **QUESTION**: why this measure?
  
  - "By multiplying this distance measure by the EU salience measure, EU issue entrepreneurship captures the extent to which a party adopts a position away from the political mainstream (that is, is more Eurosceptic) and attaches importance to this position" (p. 256)
  
  - MODEL: issue entrepreneurship on change in vote share + govt. status, l/r position  and l/r position squared (capture extremism effect), party size, country dummies and nested in years
  
  - **QUESTION:** why these control variables?

## Empirical evidence

```{r, fig.show="hold", out.height = "70%", out.width="50%", fig.cap = "Table 1 in Hobolt and De Vries (2012)"}
include_graphics(tbl1)
```

## Empirical evidence

```{r, fig.show="hold", out.height = "70%", out.width="55%", fig.cap = "Table 3 in Hobolt and De Vries (2012)"}
include_graphics(tbl3)
```

## Conclusion & debate

- Main take-away?

- What do you think of the paper?

- Future avenues of research?

- Questions or doubts?

# More recent findings

## More recent findings

- "Issue Entrepreneurship and Multiparty Competition" [@Hobolt2015]

  - Update of the previous paper
  
  - Main analysis and theory are the same
  
  - Two additions:
  
    1. Challenger parties that "hold non-mainstream policy positions, and have suffered electoral defeat are most likely to act as issue entrepreneurs" (p. 1177)
    
    2. "(...) parties only highlight a previously ignored issue when it is conceived as a potentially winning strategy. Specifically, parties only promote a previously ignored issue when their policy position is close to the mean voter position and when their position is characterized by a lack of internal dissent." (p. 1177)
    
  - Therefore, the **interaction between demand and supply side** matters

## More recent findings

- "Fleeing the centre: the rise of challenger parties in the aftermath of the euro crisis" [@Hobolt2016]

  - "voters who were economically adversely affected by the crisis punish mainstream parties both in government and in opposition by voting for challenger parties" (abstract)
  
  - "the choice of specific challenger party is shaped by preferences on three issues that directly flow from the euro crisis: EU integration, austerity and immigration" (abstract)
  
  - Important **link between the role of economic voting and crises AND the role of second-dimension issues** on altering party-systems

# Additional readings

## Additional readings

- "Party convergence and vote switching: Explaining mainstream party decline across Europe" [@Spoon2019]

  "(...) it is hypothesised that as mainstream parties converge on the left-right scale, voters will switch from supporting a mainstream party to a non-mainstream party in the next election as they look for an alternative that better represents their ideological views." (abstract)

# References

## References {.allowframebreaks}

<div id="refs"></div>
