---
title: "Factors of change I: demand-side explanations"

subtitle: "Session 10"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

navigation: horizontal 

header-includes:
- \usepackage{booktabs}

output:

  beamer_presentation:

    #includes:
    #  in_header: header_pagenrs.tex

    latex_engine: xelatex
    
    theme: "Pittsburgh"
      
    #fonttheme: "serif"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: TRUE
    
    toc: TRUE
    
    number_sections: TRUE


---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files

# Loading images' paths
fig2 <- "images/Oesch_2008_fig2.png"
```

# Explaining party system change: demand vs. supply factors

## Explaining party system change: demand vs. supply factors

- So far, we have studied party-systems from a holistic perspective:

  - Comparatively (e.g., differences between SE and NWE)
  
  - Historically (e.g., from 1920s to the 1960s)
  
  - Focusing either on general trends (e.g., Green and Far Right parties)
  
  - Or on strong specific shocks (e.g., Tangetopoli)
  
  - We have reviewed some theories of party-system change (e.g., the freezing hypothesis)
  
  - Or try to adequate empirical patterns to the existent theories (e.g., cleavage theory)

## Explaining party system change: demand vs. supply factors
  
- But what happens if we want to decompound these over-arching explanations and focus in more specific phenomena?

- From analyses at the **aggregate**-level (country, party-system or election level) to analyses at the **party and voter** (i.e., individual) levels

## Explaining party system change: demand vs. supply factors

- One way to do this analysis is to distinguish between **supply** and **demand** side explanations

- The supply side (tomorrow):

  + Level of analysis: parties
  
  + Question: how do new parties get votes?
  
- The demand side (today):

  + Level of analysis: individuals/voters
  
  + Question: who votes for new parties and why?

# Factors of change I: demand-side explanations

## Factors of change I: demand-side explanations

- Shifting from the analysis of countries, party-systems and elections to **voters** (not only: meso-level units of observations too [e.g., regions or municipalities])

- **What demand-side factors may explain voters shifting to new/challenger parties?** (think on implications of  the theories we have reviewed!)

  + Age
  
  + Gender
  
  + Education
  
  + ...
  
  + **Class**

- ***Explaining Workers’ Support for Right-Wing Populist Parties in Western Europe*** [@Oesch2008]

# The changing role of class: "Explaining Workers’ Support for Right-Wing Populist Parties in Western Europe" [@Oesch2008]

## The puzzle and research question

- The **puzzle**:

  + A emerging new kind of far right [@Ignazi1992]
  
    + Austrian Freedom Party, Belgian Flemish Block, French National Front, Danish People's Party, Norwegian Progress Party
  
    + Italian Lega Nord, Swiss Poeple's Party
  
  + **New type of working-class party?**
  
  + **Class cleavage traditionally explained vote for left/right parties**: the sociological premise is that **low-income voters should support parties that proposes a stronger intervention of the state on the economy to minimize risk exposure**
  
- Therefore, ***Research Quesion***: **why workers are more likely than any other class** (with exceptions) **to vote for RPPs?**

## Hypotheses

  + Emerging theses to explain **why workers would shift from socialist and communist parties**:
  
    H1. Economic anxieties
    
    H2. Cultural threat
    
    H3. Social alienation/political dissatisfaction
    
## Hypotheses

H1. Economic anxieties

  - Anti-immigration attitudes gained center stage
  
  - Two possible motives:
  
    + Economic reasons (fear of pressure on wages and welfare benefits)
    
    + Cultural reasons (unease about multiculturalism and the granting of citizens rights to foreigners)
    
  - **Economic reasons**
  
    + "Modernization losers" (lack of convertible skills)
    
    + Two hypotheses:
    
      - Vote for RPPs as a response to perceived competition from immigrants for jobs, wages, social benefits, etc.(*economic protectionism hypothesis*)
      
      - Vote for RPPs because of the opinion that social benefits should be restricted to nationals (*welfare competition hypothesis*)

  - **QUESTION:** how is the second hypothesis different from the first one?
    
## Hypotheses

H2. Cultural threat

  - Questions of community and identity instead of distributional conflicts
  
  - "The emergence of RPPs thus represents the authoritarian response to the dominance of libertarian values and multicultural models of living" (p. 352)
  
  - "Individuals in less privileged class locations are expected to be more receptive to the concept of “national preference” and an exclusionary notion of citizenship" (p. 352)
    
    + Two hypotheses:
    
      - Vote for RPPs because of fear that a country's culture is undermined by immigrants (*cultural protectionism hypothesis*)
      
      - Vote for RPPs because of an exclusive definition of citizenship (*differential nativism hypothesis*)

  - **QUESTION:** why should workers be more likely to hold this kind of opinions?
    
## Hypotheses

H3. Social alienation/political dissatisfaction

  - Protest voting as a cornerstone explanation (RPPs as protest parties)
  
  - "Confronted with mass unemployment and stagnating real income, workers are expected to be particularly prone to nurture resentment against the political establishment" (p. 353)

  - Alternative explanation: social anomie (remember the 'fading party ties' hypothesis?)
    
    + Two hypotheses:
    
      - Vote for RPPs because of dissatisfaction with how democracy works (*protest voting hypothesis*)
      
      - Vote for RPPs when not integrated into intermediary networks (*social disorganization hypothesis*)

  - **QUESTION:** what do the author mean by intermediary networks?

## Research design

- Five Western European countries: Austria, Belgium, France, Norway, and Switzerland

- In all five countries there is a large right-wing populist party

- "We considered only RPPs that obtained at least 5 percent of votes in parliamentary elections in or prior to 2002" (p. 354)

- **QUESTION:** which are these parties in each country?

- **QUESTION:** what do you think of this case selection strategy?

## Research design

- RPPs:

  1. a location at the extreme right on the ideological axis ranging from libertarian-universalistic to traditionalist-communitarian positions
  
  2. a populist anti-establishment discourse
  
  3. a hierarchical internal structure which sets them apart from pluralist parties

- **QUESTION:** does this definition coincide with the Mudde's definition?

## Data and models

- ESS 02/03

- ESS underestimation of PRR voting (**why?**)

- DV: self-declared vote for a PRR party

- IV: Sex, age, **class** + **one attitudinal variable for each hypothesis**

- Expectation: remove the effect of the working class when introducing the attitudinal variables

- **QUESTION:** what do you think of this empirical strategy?

## Empirical evidence

- Proportion of RPP supporters by class and country

```{r, fig.show="hold", out.height = "60%", out.width="60%", fig.cap = "Figure 2 in Oesch (2008)"}
include_graphics(fig2)
```

## Empirical evidence

- Proportion of RPP supporters by class and country (p. 356)

  1. "First, the thesis of a proletarianization of the right-wing populist parties’ electorate clearly seems correct for the five countries on which this study focuses"
  
  2. "Second, alongside the two little-privileged classes of production and service workers, a third category is over-represented among RPP followers, namely, small-business owners"
  
  3. "Third, two categories show very little sympathy for RPPs in all five countries. This applies to the traditional bourgeoisie (comprising large employers and self-employed professionals) and, above all, sociocultural professionals and semi-professionals"

## Empirical evidence

- Main conclusions from the multivariate analysis (p. 365):

  - "only one variable is significantly linked with RPP support in all five countries: people’s fear that immigrants undermine the country’s culture"
  
  - In general, cultural explanations work better than economic explanations
  
  - In Switzerland and Austria, alineation doesn't have any effect but it is very important in the other countries, **why??**
  
  - "It is noteworthy that these determinants of RPP support do not entirely pick up the class effect." **What does this mean?**
  
  - Conclusion: "voters’ attitudes make a much bigger explanatory contribution than their socio-demographic characteristics"

## Empirical evidence

- Cross-country differences:

  1. Switzerland and Austria
  
  2. Belgium, France and Norway

## Conclusion & debate

- Main take-away?

- What do you think of the paper?

- Future avenues of research?

- Questions or doubts?

## Recent findings

- "Electoral competition in Europe’s new tripolar political space: Class voting for the left, centre-right and radical right" [@Oesch2018]

- Tri-polar instead of bi-polar competition:

  1. the left
  
  2. the centre-right
  
  3. the PRR

## Recent findings

- Five class categories and two types of voting classes:
  
  1. Preserve classes: sociocultural professionals for the left and large employers and managers for the centre-right
  
  2. Contested strongholds: small business owners for the centre-right and the PRR and working-class for the left and the PRR

- Conclusion: "Old patterns are structured by an economic conflict: Production workers vote for the left and small business owners for the centre-right based on their economic attitudes. In contrast, new patterns are linked to the rise of the radical right and structured by a cultural conflict"

## Additional readings

- "The Paradox of Well-being: Do Unfavorable Socioeconomic and Sociocultural Contexts Deepen or Dampen Radical Left and Right Voting Among the Less Well-Off?" [@Rooduijn2018]

- "Education and Anti-Immigration Attitudes: Evidence from Compulsory Schooling Reforms across Western Europe" [@Cavaille2019]

# References

## References {.allowframebreaks}

<div id="refs"></div>