---
title: "Changing party-systems? Wrapping-up workshop on electoral volatility"

subtitle: "Session 07"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

navigation: horizontal 

output:

  beamer_presentation:

    #includes:
    #  in_header: header_pagenrs.tex

    latex_engine: xelatex
    
    theme: "Pittsburgh"
      
    #fonttheme: "serif"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: TRUE
    
    toc: TRUE
    
    number_sections: TRUE

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files

# Loading images' paths
me_mair <- "images/me_and_mair.png"
```


# Overview of the last session

## Overview of the last session

Take-home points from Session 6:

  1. There is **a bunch of ways of assessing party system change**, from volatility indices to compound indices of party system closure or restraint
  
  2. The **debate on party system change** has revitalized in the past years. According to some analysts, the general bias towards stability of party systems remains strong, while others suggest that some European party systems have de-institutionalized in the last decade
  
  3. There is not a correct **answer** and it **depends a lot on the way you conceptualize and measure party system change**
  
  4. There is an **open venue for new research and innovation!**

- **In this session** we will move from the theory and the review of previous analyses to **perform one empirical analysis ourselves**

# A replication exercise

## A replication exercise

- **What is replication?**

- Replication consists of the **repetition of an earlier study** using sufficiently **similar methods** and under **sufficiently similar circumstances**

- **Different types**:

  1. Systematic repetition of the exactly same analysis over the exact same data
  
  2. Additional analyses over the exact same data
  
  3. Exact same analyses over different but similar data

  4. Additional but related analyses over different but similar data

## A replication exercise

- Important for science because of:

  1. **Verifiability**
  
  2. **Validity**
  
  3. **Scope conditions** testing

# An empirical exercise with real world data

## An empirical exercise with real world data

- **Why are we doing a replication exercise ourselves?**

- First of all, just to show that **it can be done! **

- We can test the implications of what we have seen so far, validate previous results, and maybe even discover something new

- We will focus on the most used indicator to assess party system change: **electoral volatility**

- Specifically, we will use the data produced by Vincenzo Emanuele [-@Emanuele2015] on electoral volatility in Western Europe from 1945 to 2015

- This is one of the dataset that he and his colleagues use to analyze party system change [i.e., @Chiaramonte2017]

# The empirical exercise

## The empirical exercise

- First of all, **download an install R and RStudio**

- **R** is a free software environment for statistical computing and graphics with a vast collection of packages, all of them open source

- **RStudio** is an Integrated Development Environment (IDE) specifically suited for R 

- It is also free and open source!

- In addition, it supports other languages too (these slides have been done with Beamer in RStudio!) 

## The empirical exercise

- First step, **download the latest version of** [R](https://cloud.r-project.org/) **and** [RStudio](https://www.rstudio.com/products/rstudio/download/)

- Second step, go to the website of the course and **download the file replication_exercise.Rmd from the folder** [07_session](https://github.com/ACanalejo/cwps_continuity_and_change_unilu2022/tree/master/07_session)

- Third step, **open the file replication_exercise.Rmd with RStudio**

# More on replication: what you can do

## More on replication: what you can do

- Look at the papers in the course bibliography

- **The most recent papers normally come with replication files** (most high ranking journals nowadays demand these files for publication)

- If not, **they will refer to the data they use** somewhere in the paper and you can always google it! This usually works

- Now that you have R and you know the basics, **you can repeat this exercise with other datasets**

- Of course, **R programming can be difficult** at the beginning **but it is a matter of practice** and ***lot of google searching***

- **Is there anything specific that you would like to do?**

# Feedback and or organization of the second block II

## Feedback and or organization of the second block II

- **Congratulations!** We have finished the first block of the seminar

- In the second block we will move from the country/party-system level to the party and voter levels:

  1. Review some demand-side factors that explain party system change/continuity
  
  2. Review some supply-side factors that explain party system change/continuity

- You will also have to present your work

- Finally, we will review  what we have discussed together

## Feedback and or organization of the second block II

- There are three final things that we must take care of now:

  1. Sort out the presentations
  
  2. Get your feedback on the course so far
  
  3. Is there anything that you would like to discuss in the next session that is not included on the syllabus? 
  
- Would you like a different arrangement for the rest of the sessions?

## Feedback and or organization of the second block II

That's all! 

**Many thanks!** 

```{r, fig.show="hold", out.height = "60%", out.width="50%", fig.align="right"}
include_graphics(me_mair)
```

# References

## References {.allowframebreaks}

<div id="refs"></div>