---
title: "Party-system formation in Europe and the 'freezing hypothesis'"

subtitle: "Session 01"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

output:

  beamer_presentation:
    
    latex_engine: xelatex
    
    theme: "Singapore"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: TRUE
    
    toc: TRUE
    
    number_sections: FALSE

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files

# Loading images' paths

germany49 <- "images/german_elec_results_1949.png"
germany21 <- "images/german_elec_results_2021.png"

```


# 01. Introduction to the course

## 01. Introduction to the course

- Presentations

- The course: what is it about and what are we going to do

  + Review of party-system formation in WE
  
  + Analysis of main pattern of continuity and change
  
  + Discussion of the main factors explaining these patterns
  
  + Students' presentations
  
  + Wrap-up session

## 01. Introduction to the course

- Course structure and class dynamics

  + Not a lecture
  
  + Interrupt me when you feel like it
  
  + I hope to learn too!

- Today's programme

  1. Party-system formation in Europe and the 'freezing hypothesis'

  2. 'Green parties' and new social movements (70s-80s)

  3. The emergence of new 'radical right parties' (80s-90s)

  4. The 'freezing hypothesis' revisited: Mair & Bartolini in the 90s

# 02. Mapping the terrain: political parties and party-systems

## 02. Mapping the terrain: political parties and party-systems

- What is a political party?
  
- A political party, for the sake of this seminar, is a hierarchical organizations that compete for votes to achieve office, with the aim of implementing its preferred policies

- What is a party-system?
  
- The interaction of parties competing in the given electoral market. It is characterised by the number, size, and ideological preferences of the parties, among other aspects  


## 02. Mapping the terrain: political parties and party-systems

- Why is it important to study political parties and party-system?

- It help us to understand a myriad of political phenomena. Each party-system indicates levels of societal fragmentation and polarization, the most likely electoral and policy outcomes at a given moment in time, potential coalitions, etc. Comparatively, it helps us to assess dynamics of evolution of the public opinion, foresee potential reforms, alliances...

- In the end, party politics is a crucial aspect of political science: 'the science of (public) power'

- In this seminar, we will try to understand why European party-systems looks like they are today, considering how they were born and evolve, which will surely help us to assess the future that awaits us (but never predict it!)

# 03. The origins of party-systems in Europe

## 03. The origins of party-systems

```{r, figures-side, fig.show="hold", out.height = "70%", out.width="40%", fig.cap = "German Federal electoral results, images from Wikipedia"}
par(mar = c(6, 6, 2, 2))
include_graphics(germany49)
include_graphics(germany21)
```

## 03. The origins of party-systems

- How were political parties born?

- 

# 04. European party-systems: similitudes and differences

## 04. European party-systems: similitudes and differences

# 05. The 'freezing hypothesis': continuity from the 1920's to the 1960's

## 05. The 'freezing hypothesis': continuity from the 1920's to the 1960's
