---
title: "Party-system formation in Europe and the 'freezing hypothesis'"

subtitle: "Session 01"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

navigation: horizontal 

output:

  beamer_presentation:
    
    #includes:
    #  in_header: header_pagenrs.tex
    
    latex_engine: xelatex
    
    theme: "Pittsburgh"
      
    #fonttheme: "serif"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: TRUE
    
    toc: TRUE
    
    number_sections: TRUE
    
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files

# Loading images' paths
germany49 <- "images/german_elec_results_1949.png"
germany21 <- "images/german_elec_results_2021.png"

```


# Introduction to the course

## Introduction to the course

- Presentations

- The course: what is it about and what are we going to do

  + Review of party-system formation/evolution in WE
  
  + Analysis of the main patterns of continuity and change
  
  + Discussion of the main factors explaining these patterns
  
  + Students' presentations
  
  + Wrap-up session

## Introduction to the course

- Course structure and class dynamics

  + Not a lecture, I appraise debate
  
  + Interrupt me whenever you want
  
  + I hope to learn too!

- Today's programme

  1. Party-system formation in Europe and the 'freezing hypothesis'

  2. Green parties and the silent revolution

  3. New far right parties and the counter-revolution

  4. Economic voting and the Great Recession: new politics in Southern Europe

# Mapping the terrain: political parties and party-systems

## Mapping the terrain: political parties and party-systems

- What is a political party?
  
    + A political party, for the sake of this seminar, is a hierarchical organization that compete for votes to achieve office with the aim of implementing its preferred policies

- What is a party-system?
  
    + The interaction of parties competing in the given electoral market. It is characterised by the number, size, and ideological preferences of the parties, among other aspects  


## Mapping the terrain: political parties and party-systems

- Why is it important to study political parties and party-systems?

    + It help us to understand a myriad of political phenomena. Each party-system indicates levels of societal fragmentation and polarization, the most likely electoral and policy outcomes at a given moment in time, potential coalitions, etc. Comparatively, it helps us to assess dynamics of evolution of the public opinion, foresee potential reforms, alliances...

- In the end, party politics is a crucial aspect of political science: 'the science of (public) power'

- In this seminar, we will try to understand why European party-systems looks like they are today, considering how they were born and evolved, which surely help us to assess future events (although never fully predict them)

# The origins of party-systems in Europe

## The origins of party-systems in Europe

- Here we don't care much about the origins of political parties themselves ([some say that they were born in the UK in the early 18th century...](https://en.wikipedia.org/wiki/Early-18th-century_Whig_plots)), but on **the determinants of party-systems**. Two main approaches can be highlighted [@Boix2007]:

  1. Institutional approaches

  2. The **socio-historical approach**, famously indebted to Lipset and Rokkan [-@Lipset1967a]

## The origins of party-systems in Europe

1. Institutional approaches

    + Institutions (the rule of the game) determine the number, size and characteristics of competing parties via direct and indirect (the so-called psychological) effects
  
    + For example, the **Duverger law** estipulates that single-ballot plurality-rule elections (such as first-past-the-post system) structured within single-member districts tend to favor a two-party system
  
    + Effects can be **direct** when institutions automatically shape the conversion of behaviour (i.e., vote) into outcomes (i.e., parlimentary seats)
  
    + Effects can also be **indirect** when people modify their behavior prior to the application of the rule because they anticipate direct effects (e.g., I anticipate that my preferred party may not win a seat due to the majoritarian electoral rule, therefore I cast a ballot for my second preferred option, which has higher probabilities to win a seat)  

    + **Any other example** of direct vs. indirect effect?

## The origins of party-systems in Europe

2. The **socio-historical approach**

    + The structure of party-systems in Europe is mainly due to the lines of conflict politicized in each country
  
    + These lines of conflict, that we call **cleavages**, are delved into the formation of each nation-state and also fostered by common economic and cultural shocks
  
    + These common shocks affect each country differently nonetheless (most importantly to Lipset and Rokkan: the industrial revolution)
  
    + A cleavage is characterised by 1) a *social divide* of which citizens at each side develop 2) *social consciousness* and turn into 3) *political demands* 
  
    + Any example of what may be and what may not be a cleavage nowadays in any European country?

## The origins of party-systems in Europe

2. **The socio-historical approach**

    - Lipset & Rokkan establish four cleavages:
  
      1. **Class** cleavage (owner versus worker)
    
      2. **Urban-rural** cleavage (industry versus land)
    
      3. **Centre-periphery** cleavage (dominant versus subject culture)
    
      4. **State-church** cleavage (state versus church institutions)
    
    - The first two cleavages would be directly influenced by social changes brought by the industrial (and liberal) revolutions
  
    - The other two would be influenced by each national revolution, and it would depend on specific preconditions (e.g., strong privilege of the Catholic Church in Italy / strong minority elites in Spain)

## The origins of party-systems

- The transformations of cleavages into party-systems depends on the specific **political opportunity structure**\footnotemark, that determines how far within a *sequence of thresholds* any movement pressing forward new sets of demands arrives

  1. The threshold of *legitimation* (i.e., opposition recognition)

  2. The threshold of *incorporation* (i.e., participation rights)

  3. The threshold of *representation* (i.e., institutional presence)

  4. The threshold of *majority power* (i.e., participation in government)
      
        \footnotemark[1]Lipset and Rokkan mention (1) the traditions of decision-making in the polity, (2) the channels of expression and mobilization of protest, **(3) the opportunities, the payoffs and the costs of alliances**, (4) and the possibilities, the implications and the limitations of majority rule in the system 

# European party-systems compared

## European party-systems compared

The basic party families\footnotemark:

  + *Socialist and communist parties*: competition based on the class cleavage
  
  + *Agrarian parties*: competition based on the urban-rural cleavage
  
  + *Regionalist/independentist parties*: competition based on the centre-periphery cleavage
  
  + *Religious and people's parties*, such as the Christian democrats: competition based on the state-church cleavage
  
Beyond this simplification, most parties compete on several overlapped arenas (a clearly example is the *Liberals party family*; class and urban-rural), but depending on the most salient issues, some parties will be stronger than others

  \footnotemark[2]Based on Camia and Caramani [-@Camia2012]
  
## European party-systems compared

- The main cleavage structuring party competition in Europe is the class cleavage

- The class cleavage is not only the one that accounts for the most part of variation on voting behavior across social groups

- It is also the only cleavage that holds across European countries and over time

- Depending on the combination of parties and their modes of interaction, we can talk of two main types of party systems [@Sartori2005]:

    1. **Moderate party systems** have relative small ideological distance between the major parties, a propensity to form coalitions between parties of different views and predominantly centripetal competition
    
    2. **Polarised party systems** have two main strongly opposed party blocks, presence of small 'anti-system' parties and substantial ideological distance between parties at each side, with predominantly centrifugal competition
    

## European party-systems compared

- The existence of **overlapping/coinciding** vs. **cross-cutting cleavages** raises different patterns fo party competition

- For exmaple, **on the strength of Socialist/Labor parties**\footnotemark:

  + Protestant countries with the smoothest nation-building experience (e.g., Britain, Denmark, and Sweden) produced the most cohesive and integrated Socialist parties

  + Catholic countries with recent or difficult national revolutions had deeply and alienated working class movements instead (e.g., France, Spain, and Italy)

  + The countries in between has the additional factor or whether the nation-building elites were friendly with the Catholic Church during the period of mass education and enfranchisement (e.g., Austria, Ireland and Belgium)

- What are the mechanisms underlying these processes?

  \footnotemark[3]Based on Merkl [-@Merkl1969]

## European party-systems compared

- Another example, on the factors facilitating the emergence of agrarian parties\footnotemark (e.g., prevalent in Nordic countries vs. Southern Europe):

  1. Weak industrial-urban centers at the time of suffrage extension
  
  2. Dominant family-size farms or legally protected lease-holding
  
  3. Cultural and economic barriers between countryside and cities
  
  4. Insignificance of Catholicism

- There are typologies beyond the scope of this seminar that help to explain differences within clusters (e.g., consensual vs. majoritarian institutions)

\footnotemark[4]Based on Merkl [-@Merkl1969]

# The 'freezing hypothesis'

## The 'freezing hypothesis': the German case

```{r, figures-side, fig.show="hold", out.height = "70%", out.width="40%", fig.cap = "German Federal election results, images from Wikipedia"}
par(mar = c(6, 6, 2, 2))
include_graphics(germany49)
include_graphics(germany21)
```

## The 'freezing hypothesis': continuity between the 1920s and the 1960s

- "The authors conclude (...) that the current European party systems, with some exceptions, reflect the survival of the 'cleavage structures' of the 1920's in spite of the onslaught of Fascism and National Socialism and World War II" [@Merkl1969]

- Lipset and Rokkan highlight the shockingly pervasive stability of European party-systems despite intermittent wars, autocratic periods and revolutions

- But are European party-systems currently frozen? What does the case of Germany suggest to you?

## The 'freezing hypothesis': let's debate!

- ***Final activity!*** Let's divide the class in two groups. You have 5-10 minutes to discuss potential explanations against or in favour of the freezing hypothesis, that is, why should we expect party-systems to remain 'frozen' until nowadays (or not). Someone from each group should summarise the main points and then we will briefly debate. 

- The activity begins now!

# References

## References {.allowframebreaks}

<div id="refs"></div>