---
title: "Wrapping-up session: review, questions and doubts"

subtitle: "Session 08"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

navigation: horizontal 

header-includes:
- \usepackage{booktabs}

output:

  beamer_presentation:

    #includes:
    #  in_header: header_pagenrs.tex

    latex_engine: xelatex
    
    theme: "Pittsburgh"
      
    #fonttheme: "serif"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: TRUE
    
    toc: FALSE
    
    number_sections: TRUE


---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files

# Loading images' paths
```

# Welcome back!

## Welcome back!

- Recent electoral news in Europe

  + The Hungarian parliamentary elections
  
    - On Sunday 3rd April
  
    - "We won a victory so big that you can see it from the moon, and you can certainly see it from Brussels" (Viktor Orbán)

    - The 199 members of the National Assembly are to be elected by two methods; 106 are elected in single-member constituencies by first-past-the-post voting, with the remaining 93 elected from a single nationwide constituency mostly by proportional representation

    - The electoral threshold is set at 5%, although this is raised to 10% for coalitions of two parties and 15% for coalitions of three or more parties

## Welcome back!

- Recent electoral news in Europe
    
  + The French presidential elections
  
    - On Sunday 10th April
    
    - The polls: https://www.politico.eu/europe-poll-of-polls/france/

    - What does this say about party system change and the content reviewed in our course?

# Block I review

## Block I review: doubts and questions

- Doubts and questions!

  + Things more problematic

  + Things more interesting
  
  + Other?

## Block I review: a summary

- Party-system formation:

  - Session 1: **cleavage theory**

    + The three components of cleavages: social and political features linked by an identity component
  
    + Overlapping/cumulative vs. cross-cutting cleavages
  
    + Why does the cleavage structure determines the features of party systems and party competition?
  
## Block I review: a summary

- Party-system change: a chronological review in WE

  - Session 2: **Green parties**

    + The silent revolution (value change hypothesis)
  
    + New Social Movements
  
    + Asymmetrical success of new parties: strategies and insitutions?

## Block I review: a summary

- Party-system change: a chronological review in WE

  - Session 2: **Green parties**

```{r tbl, echo = FALSE, message = FALSE, warning = FALSE}


country <- c("Germany", "France")

elec_rules <- c("Majoritarian",
                "Proportional")

unions_left <- c("Medium",
         "Strong")

left_segm <- c("No",
         "Yes")

left_govt <- c("No",
         "No")

fed_inst <- c("Yes",
         "No")

type_green <- c("New Left",
         "Conservative")

av_green_vote_78_91 <- c("5.1%",
         "1.2%")


tbl <- data.frame(country,
                  elec_rules,
                  unions_left,
                  left_segm,
                  left_govt,
                  fed_inst,
                  type_green,
                  av_green_vote_78_91)

tbl <- t(tbl)

rownames(tbl) <- c("Country",
                    "Electoral rules",
                    "Union's control over the left",
                    "Segmentation of the Left",
                    "Left in Government",
                   "Federal institutions",
                    "Type of Green challenger",
                    "Average Green vote 78-91")

tbl %>% 
  kable(format = "latex", 
        booktabs = T,
        digits = 2, 
        align = "c",
        caption = "Early success of Green Parties in Germany and France") %>% 
row_spec(1, bold = TRUE)
```

## Block I review: a summary

- Party-system change: a chronological review in WE

  - Session 3: **new far-right parties/PRR parties**

    + Counter-reaction from the right
  
    + Fostered by economic and cultural changes
  
    + Top-down vs. bottom-up mobilization
  
## Block I review: a summary

- Party-system change: a chronological review in WE

  - Session 4: **economic voting** and crises

    + What is a crisis?
  
    + Economic voting theory
  
    + Rapid party system change due to valence (vs. positional) issues
  
## Block I review: a summary

- Party-system change: the theory

  - Session 5: **cleavage theory revisited**

    + Dimensional models of political behavior and party competition
  
    + Issue saliency
  
    + A new cultural cleavage
  
## Block I review: a summary

- Party-system change: the empirics

  - Session 6: **empirical assessment of party system change**

    + electoral volatility
  
    + other measures?
  
    + general bias towards stability?
  
# Block II structure

## Block II structure

- Friday 08.04.2022 (09:00 – 17:00)

  + **Session 8** (09:00 – 10:30): **Wrapping-up session: review, questions and doubts**

  + **Session 9** (10:45 – 12:15): **Students presentations I**
  
    - Andrea: The Netherlands
    
    - Miguel: Portugal

  + **Session 10** (13:45 – 15:15): **Factors of change I: demand-side explanations**
  
    - "Explaining Workers’ Support for Right-Wing Populist Parties in Western Europe: Evidence from Austria, Belgium, France, Norway, and Switzerland" [@Oesch2008]

  + **Session 11** (15:30 – 17:00): **Students presentations II**
  
    - Rèmy: France
    
    - Luca: Norway
    
## Block II structure

- Saturday 09.04.2022 (09:00 – 15:15)

  + **Session 12** (09:00 – 10:30): **Factors of change II: supply-side explanations**
  
    - "When dimensions collide: The electoral success of issue entrepreneurs" [@Hobolt2012]

  + **Session 13** (10:45 – 12:15): **Students presentations III**
  
    - Judith: Germany

    - Álvaro: Greece

  + **Session 14** (13:45 – 15:15): **Wrapping-up discussion, advanced topics and feedback**

## Block II structure

- Any other doubts or question?

- Thanks!

# References

## References {.allowframebreaks}

<div id="refs"></div>
