---
title: "The cultural dimension: cleavage theory revisited"

subtitle: "Session 05"
  
author: "Álvaro Canalejo-Molero"

date: "Spring Term 2021-2022"

institute: "University of Lucerne"

bibliography: ../00_bibliography/cweps_syllabus.bib

csl: ../00_bibliography/apa.csl

link-citations: yes

linkcolor: blue 

fontsize: 10pt

linestretch: 2

navigation: horizontal 

output:

  beamer_presentation:

    #includes:
    #  in_header: header_pagenrs.tex

    latex_engine: xelatex
    
    theme: "Pittsburgh"
      
    #fonttheme: "serif"
    
    fonttheme: "structurebold"
    
    slide_level: 2
    
    incremental: FALSE
    
    toc: TRUE
    
    number_sections: TRUE

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# Loading libraries
library(knitr) # For knitting document and include_graphics function
library(kableExtra) # for LaTeX tables in Rmarkdown
library(png) # For loading and grabbing the dimensions of png files
library(tidyverse) # For loading the tidyverse environment
library(ggpubr) # For arranging ggplot plots

# Loading images' paths
aus2006_gre2007 <- "images/aus2006_gre2007.png"
```


# Overview of the last sessions

## Overview of the last sessions

Take-home points from Session 1 to 4 (**who's up for a summary?**):

1. Session 1 

    - European **party systems are rooted on the social transformations that led to state formation**. They reflect the structure of conflict within each country and are shaped by reinforcing political institutions. From the 1920s to the 1960s, they remained mostly unchanged 

2. Session 2

    - **Since the 60s**, a transformation in the economic structure and generational replacement led to a **shift from materialistic to post-materialistic values**. This provided the necessary context to initiate some changes in the configuration of party systems, **mainly driven by green and new left parties**

## Overview of the last sessions

Take-home points from Session 1 to 4:

3. Session 3

    - **As a counter-reaction** to the changes initiated in the 60s, a wave of **new far right parties** took place, introducing a **progressive transformation on the** traditional class cleavage and the overall **conflict structure**

4. Session 4
  
    - In contrast to the progressive cumulative change in NWE, some **SE countries**, that had been less affected by the green and PRR wave, **experienced a series of party systems shock after political and economic crises**

# Introduction to the spatial model I: the unidimensional model

## Introduction to the spatial model I: the unidimensional model

- **But how can we make sense of these changes?**

- In this session we will cover the most recent **updates of the cleavage theory**, that account for the transformation of western European party systems 

- To follow these updates, we will first review some basic notions of the **spatial theory of voting and party competition**

## Introduction to the spatial model I: the unidimensional model

- The spatial (or economic) theory of voting and party competition was proposed by **Downs** [-@Downs1957]

- He proposed an analysis of elections as **electoral markets** in which the **demand side** was the electorate and the **supply side** was the competing parties

- In his initial **model**, based on the USA context, two parties competed to **maximize voters** over **one dimension**

- We will use this basic set-up for simplicity, although **Europe** is mostly characterize by **multi-party systems**

## Introduction to the spatial model I: the unidimensional model

- So first of all, what do we mean by a **dimension**?

  1. Mathematically -> a vector that represents a a set of preferences over a continuum

  2. In our setup -> policy preferences that range from 0 (the left) to 10 (the right)

\

\begin{beamercolorbox}[center]{title}
the left 0 --- 1 --- 2 --- 3 --- 4 --- 5 --- 6 --- 7 --- 8 --- 9 --- 10 the right
  \end{beamercolorbox}

## Introduction to the spatial model I: the unidimensional model

- To recall:

    1. Models are simplifications of reality
  
    2. "All models are wrong, but some are useful" (G. Box)
  
    3. The key for interpreting any model: understanding the assumptions

- The **main assumptions of the Downsian model** are that parties aim to **maximize votes** to win the election 

- The second assumption is that each voter have a fixed policy preference over the continuum (**preference distribution**) and **will vote to the closest party if there is one closest alternative**

## Introduction to the spatial model I: the unidimensional model

- Each party depart from a different side of the spectrum (representing **each side of a cleavage**)

- Then, each political party takes the **policy stance** (i.e., position over the continuum) that maximizes its chances of winning the election by **approximating the preference of the median voter** (i.e., the preference that leaves an equal number of voters at each side of the distribution)

- Let's imagine that in our case we have two parties, each of them representing each side of the left-right ideological dimension, that want to maximize their chances of winning the next election by moderating their positions

  1. The first party is a **centre-left party**, that takes a value of **4**

  2. The second party is a **centre-right party**, that takes a value of **6**

  3. **Preferences are normally distributed** within the population

## Introduction to the spatial model I: the unidimensional model

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7,fig.height=4, fig.cap ="Unidimensional space with two parties and abstention"}
# Simulated competition models

## L-r preferences
set.seed(124)

preferences_lr <- rbinom(n = 1000, size = 10, prob = 0.5)

## Orthogonal gal-tan preferences
set.seed(431)

preferences_galtan <- rbinom(n = 1000, size = 10, prob = 0.5)

## Non-orthogonal gal-tan preferences
preferences_galtan_corr <- (
  # correlated with l-r preferences (more to the left in galtan)
  0.4*preferences_lr +
  # + uncorrelated (random) error                          
  0.6*preferences_galtan)

## Creating the dataset
df <- data.frame(preferences_lr,
                      preferences_galtan,
                      preferences_galtan_corr)

#summary(df$preferences_galtan_corr)

# Manipulating the data
df <- df %>%
  mutate(centerLeft_lr = 4, # center-left party position equal on both dimensions
         centerRight_lr = 6, # center-right party position equal on both dimensions
         # preferences in unidimensional competition (median voter theorem)
         party_pref_unidim = ifelse(preferences_lr <= centerLeft_lr, 1,
                                   ifelse(preferences_lr >= centerRight_lr, 2, 0)),
         party_pref_unidim = factor(party_pref_unidim,
                                      levels = c("0",
                                                 "1",
                                                 "2"),
                                      labels = c("Abstention",
                                                 "Center-left",
                                                 "Center-right")),
        # preferences in bidimensional competition I: orthogonal gal-tan
         party_pref_bidim = ifelse(preferences_lr <= 5 &
                                 preferences_galtan <= 5, 1,
                                ifelse(preferences_lr >= 5 &
                                       preferences_galtan >= 5, 2, 0)),
         party_pref_bidim = factor(party_pref_bidim,
                                      levels = c("0",
                                                 "1",
                                                 "2"),
                                      labels = c("Abstention",
                                                 "Center-left",
                                                 "Center-right")),
         # preferences in bidimensional competition II: non-orthogonal gal-tan
         party_pref_bidim_corr = ifelse(preferences_lr <= 5 &
                                 preferences_galtan_corr <= 5, 1,
                                ifelse(preferences_lr >= 5 &
                                       preferences_galtan_corr >= 5, 2, 0)),
         party_pref_bidim_corr = factor(party_pref_bidim_corr,
                                      levels = c("0",
                                                 "1",
                                                 "2"),
                                      labels = c("Abstention",
                                                 "Center-left",
                                                 "Center-right")))

#glimpse(df)

## Visualizing unidimensional competition = bidimensional with perfectly correlated preferences
ggplot(df, aes(x = preferences_lr, fill = party_pref_unidim)) +
  geom_histogram(bins=10) +
  xlab("Preferences first dimension") +
  ylab("Frequency") +
  guides(fill=guide_legend(title="Party preference"))

```

## Introduction to the spatial model I: the unidimensional model

- Let's play with this set-up a little!

  1. What happens if we add a **centre-party** to the model?
  
  2. And if we add a **extreme-right party**?
  
  3. And if we add a **extreme-right and extreme-left party**?
  
  4. What happened if the distribution is slightly **skewed to the right** (i.e., more voters in the left)?
  
  5. And **to the left** (i.e., more voters in the right)?
  
  6. What happens if **extreme voters are apathetic**?
  
  7. What happens if the **centre-right party is in government and the economy is doing good**?
  
  8. And if **the economy is doing bad**?
  
  9. What happens if the position of the parties is **3 and 7** respectively and they have **programmatic constraints/reputation** (i.e., they cannot credibly shift without a cost)?

## Introduction to the spatial model I: the unidimensional model

- In sum, **the Downsian model**, as well as models in general, are **a tool that help us understand the reality**, by providing means to summarise (theoretical) expectations graphically and mathematically

- **Representing cleavages with dimensions of competition** help us to **analyse the interaction between parties (i.e., party systems)** as well as **parties and voters (i.e., electoral competition)**, under a given set of conditions

# Introduction to the spatial model II: the bidimensional model

## Introduction to the spatial model II: the bidimensional model

- So far, however, we have only discussed a situation in where two parties compete over a single dimension

- However, we know that **cleavages can overlap or cross-cut each other**

- If they don't completely overlap, we can interpret the **cross-cutting cleavage as a second dimension**

- Let's say, for example, that two parties compete in **Catalonia (Spain)** where there are two important dimensions

  1. The **left-right economic dimension** is rooted in the **class cleavage**
  
  2. The **pro-independence - anti-independence** is rooted in the **centre-periphery cleavage**

## Introduction to the spatial model II: the bidimensional model

- We have **two hypothetical scenarios**:

  1. In **Figure 2** the two dimensions are **partially coincidental/overlapping** 
  
      + This could be the case if the second dimension was **not at all salient/divisive**
      
  2. In **Figure 3** the two dimensions are **completely independent** (i.e., orthogonal in mathematical terms)
  
      + This could be the case if the second dimension was **extremely salient/divisive**

## Introduction to the spatial model II: the bidimensional model

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7,fig.height=4, fig.cap ="Bimensional competition with two parties and correlated preferences"}

## Visualizing bidimensional competition with correlated preferences
ggplot(df, aes(x = preferences_lr, y = preferences_galtan_corr, colour = party_pref_bidim_corr)) +
  geom_jitter() +
     xlim(0, 10) +
  ylim(0,10) +
  xlab("Preferences first dimension") +
  ylab("Preferences second dimension") +
  guides(colour=guide_legend(title="Party preference"))

```

## Introduction to the spatial model II: the bidimensional model

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7,fig.height=4, fig.cap ="Bimensional competition with two parties and orthogonal preferences"}

## Visualizing bidimensional competition with orthogonal preferences
ggplot(df, aes(x = preferences_lr, y = preferences_galtan, colour = party_pref_bidim)) +
  geom_jitter() +
   xlim(0, 10) +
  ylim(0,10) +
  xlab("Preferences first dimension") +
  ylab("Preferences second dimension") +
  guides(colour=guide_legend(title="Party preference"))

```

## Introduction to the spatial model II: the bidimensional model

This lead to a series of implications, most importantly:

  1. As the **dimensions** become **more cross-cutting**, the **space opened for new parties** becomes **larger**
  
  2. The **larger** becomes the **space opened for new parties**, the **party system** will tend to be **more fragmented**
  
  3. As **more dimensions** cross-cut each other, the **space opened for new parties** becomes **larger** and the **party system** will tend to be **more fragmented**

# The emergence of a new value-based cleavage? Introducing the cultural dimension

## The emergence of a new value-based cleavage? Introducing the cultural dimension

- In previous sessions we have discussed the importance of the **class-cleavage**

- This is the **main cleavage structuring the political competition Western Europe**

- In fact, the **left-right (or first order) dimension**, which is the most used dimension in European political studies, captures preferences **deeply rooted in the class cleavage**

## The emergence of a new value-based cleavage? The cultural dimension

- The rest of cleavages discussed by Lipset and Rokkan [-@Lipset1967a] have country-specific manifestations and, with some exceptions, have weakened with time

  1. The **state-church cleavage** has eroded as a result of growing secularization
  
  2. The **urban-rural cleavage** (although re-emerging according to some scholars) have eroded with massive migration from the country side to the city
  
  3. The **centre-periphery cleavage** is the only exception, since it has become even stronger in some cases (e.g., Catalonia in Spain or Scotland in the UK)

## The emergence of a new value-based cleavage? The cultural dimension
  
- None of them, however, explains the rise of new left and new right parties

- This has led scholars to **two alternative conclusions**:

  1. Either **cleavages have lost their explanatory power** and political competition dynamics are now driven by issue voting, short-term strategies and opportunistic/charismatic leadership
  
  2. Or a **new cleavage** is structuring the political conflict in Europe
  
- This is why scholars situated on the latter tradition argue that **Western Europe is increasingly structured by a two-dimensional political space**\footnotemark

    \footnotemark[1]And the reason why I wanted to be sure that you understand the party-system implications of spatial models with two dimensions
  
## The emergence of a new value-based cleavage? The cultural dimension

- There is a series of names that different scholars have given to the new cleavage:

  1. Integration-demarcation [e.g., @Kriesi1998]
  
  2. Universalistic-communitarian [e.g., @Bornschier2010]
  
  3. Green/alternative/libertarian(**GAL**)-Tradition-authority-national(**TAN**) [e.g., @Hooghe2018b]
  
  4. Winners-losers (of globalization) [e.g., @Kriesi2006, @Hutter2019]
  
- But all of these accounts coincide on some **basic features**

## The emergence of a new value-based cleavage? The cultural dimension

- ***Basic features*** of the new cleavage:

  1. The cleavage is rooted on  (1) the expansion of **educational rights**, (2) the transition **from an industrial to a post-industrial society** and (3) the process of **globalization**
  
  2. These three processes have generated groups of **winners** and **losers**, i.e. people that have benefited or been harmed as a result, that constitute the two sides of the conflict
  
  3. The cleavage raises opposed **cultural values and preferences**. Losers tend to manifest traditional, nativist an authoritarian-oriented values while winners tend to support progressive, multiculturalist and libertarian policies
  
  4. The parties that better represent each of the poles of this conflict are the **new left parties** for the winners and the **PRR** parties for the losers
  
# Cross-country variation and current trends

## Cross-country variation and current trends

- However, this second-dimension is **not completely orthogonal** to the first dimension

- Also, **traditional parties** (i.e., mainstream or dominant) have sometimes been **dominant on the second-dimension** too (e.g., Social Democratic parties that accommodated the demands of NSM)

- There are important regional differences:

  1. In ***NWE***, the second and first dimensions are **more orthogonal** and **new parties** have been the main drivers of shifts to the second dimension
  
  2. In ***SE***, the second and first dimensions are **less orthogonal** and **traditional** parties more frequently *own* second dimension issues

## Cross-country variation and current trends
  
- **Why these differences?**

  1. Authoritarian legacies
  
  2. Party system formation during the silent revolution (i.e., more permeable party systems)
  
  3. Weaker economic development
  
  4. Lower levels of immigration
  
  5. Already segmented party systems that absorbed the new demands
  
  6. **Others?**
  
- The **recent** irruption of PRR parties in Portugal and Spain suggest some **degree of convergence**

## The emergence of a new value-based cleavage? Thecultural dimension

- **Final activity!** Let look at two paradigmatic cases visualized with real data from 15 years ago: **what can we say about them? What do they tell about recent events?**

```{r, fig.show="hold", out.height = "50%", out.width="70%", fig.cap = "Austria 2006 and Greece 2007 compared / Source: Bornschier (2010) and Hutter and Kriesi (2019) respectively"}
include_graphics(aus2006_gre2007)
```  

# References

## References {.allowframebreaks}

<div id="refs"></div>
